Dynamic Modeling, Parameter Estimation and Uncertainty Analysis in \proglang{R}

### 2016-09-27 article.tex
Bibfile und erste Zitate

%% include your article here, just as usual
%% Note that you should use the \pkg{}, \proglang{} and \code{} commands.
\section{Introduction}
	Dynamic models are found in several research fields, ranging from physics to biology or finance. In all these fields, models are used to construct a link between theoretical concepts and empirical evidence. In a dynamic model, the single terms correspond to mechanisms or processes which are combined to a whole, giving rise to a complex system. If the single terms are associated to parameters, parameter estimation can identify those processes of the complex system that are crucial to explain the observation. In that sense, parameter estimation can be employed as an instrument to understand complex systems. Once, the link between observation and model is established, questions about the identifiability of parameters arise. The parameter space needs to be explored to see whether the estimate is unique and to determine confidence bounds.

	Although the problem of parameter estimation in non-linear dynamic systems is highly relevant and at the heart of statistical computing, up to today there are only two \proglang{R} packages published on the topic on the Comprehensive \proglang{R} Archive Network (CRAN): \pkg{FME} [REF] and \pkg{nlmeODE} [REF]. Both packages are built upon the \pkg{deSolve} package.

	The key requirement for parameter estimation is accurate and robust non-linear optimization. Many of the standard optimization routines implemented in \proglang{R} use derivative information. Indeed, if the objective function is sufficiently smooth, derivatives can considerably reduce the number of iterations and convergence is more reliable [REF]. However, the computation of derivatives in the context of ODE models holds some pit-falls [REF] which are e.g.~related to the step-size control of the integrator.

@@

There are several ways to define the differential equations. An \code{eqnvec} of equations can be explicitly formulated, analogously to a \code{c()} command. Especially for chemical reactions it can be tideous to keep track of all gain and loss terms. Therefore, \pkg{dMod} provides the \code{eqnlist} class which encodes the ODE by a list of (1) state names, (2) rate expressions, (3) compartment volumes and (4) the stoichiometric matrix. Each reaction flux occurs only once and the gain and loss is represented by the coefficients in the stoichiometric matrix [REF]. \pkg{dMod} supports the read-in of a csv file with the stoichiometric matrix and rate expressions and provides a function \code{addReaction()} to construct an \code{eqnlist} object step-by-step or add further reactions to an already existing \code{eqnlist}. The conversion to the (differential) equations is done by \code{as.eqnvec()} which does all the bookkeeping of gain and loss terms and volume ratios due to compartment transitions.

@@

>>>>

%% include your article here, just as usual
%% Note that you should use the \pkg{}, \proglang{} and \code{} commands.
\section{Introduction}
Dynamic models are found in several research fields, ranging from physics to biology or finance. In all these fields, models are used to construct a link between theoretical concepts and empirical evidence. In a dynamic model, the single terms correspond to mechanisms or processes which are combined to a whole, giving rise to a complex system. If the single terms are associated to parameters, parameter estimation can identify those processes which are crucial to explain the observation. In that sense, parameter estimation can be employed as an instrument to \textit{understand} complex systems. Once, the link between observation and model is established, questions about the identifiability of parameters arise. The parameter space needs to be explored to analyze whether the estimate is unique and to determine confidence bounds.

Although the problem of parameter estimation in non-linear dynamic systems is highly relevant and at the heart of statistical computing, to this day there are not more than two \proglang{R} packages published on the topic on the Comprehensive \proglang{R} Archive Network (CRAN), namely \pkg{FME} \citep{FME} and \pkg{nlmeODE} \citep{nlmeODE}. Both packages are built upon the \pkg{deSolve} package \citep{deSolve}.

The key requirement for parameter estimation is accurate and robust non-linear optimization. Many of the standard optimization routines implemented in \proglang{R} use derivative information. Indeed, if the objective function is sufficiently smooth, derivatives can considerably reduce the number of iterations and convergence is more reliable \citep{raue2013lessons}. However, the computation of derivatives in the context of ODE models holds some pit-falls which are e.g.~related to the step-size control of the integrator.

@@

There are several ways to define the differential equations. An \code{eqnvec} of equations can be explicitly formulated, analogously to a \code{c()} command. Especially for chemical reactions it can be tideous to keep track of all gain and loss terms. Therefore, \pkg{dMod} provides the \code{eqnlist} class which encodes the ODE by a list of (1) state names, (2) rate expressions, (3) compartment volumes and (4) the stoichiometric matrix. Each reaction flux occurs only once and the gain and loss is represented by the coefficients in the stoichiometric matrix. \pkg{dMod} supports the read-in of a csv file with the stoichiometric matrix and rate expressions and provides a function \code{addReaction()} to construct an \code{eqnlist} object step-by-step or add further reactions to an already existing \code{eqnlist}. The conversion to the (differential) equations is done by \code{as.eqnvec()} which does all the bookkeeping of gain and loss terms and volume ratios due to compartment transitions.

@@

\bibliography{mybib}{}
